Think deep for reef resilience

### 2025-08-12 paper/paper_draft.qmd
Adjust GCM description for brevity and flow

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled
1km^2^ resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4 and are a part of the Sixth Phase of the
Coupled Model Intercomparison Project (CMIP6) based on evaluation conducted by Grose et al.
[-@grose2023]. These GCMs were found to rank highly in the recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. [-@grose2023] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region. An overview of the selected GCMs are provided in @tbl-gcms. The process
described here was conducted for all 2429 reefs to create DHW projections spanning from 2025
to 2075. Code used to calculate DHW data and perform analyses involved in this investigation
can be found at **insert repo citation**.

The five selected GCMs span a range of Equilibrium Climate Sensitivity (ECS) values within
the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC defines
ECS as the long term warming caused by a doubling of carbon dioxide above its pre-industrial
concentration [@forster2021]. GCM development institutions regularly share aspects of model
development. Brunner et al. [-@brunner2020] assessed the independence of 33 models,
quantitatively classifying models into 'model development families' based on patterns in
atmospheric temperature and sea level pressure. By selecting five GCMs for investigation
that span a range of four 'model development families' [@brunner2020] (note: we assume
GFDL-CM4 aligns with GFDL-ESM4; ACCESS-ESM1-5 and ACCESS-CM2 occupy the same family), we are
able to assess climate projections over a broad range of independently derived futures
[@abramowitz2019]. The selected GCMs therefore allow for a range of possible climate futures
to be explored, capturing the uncertainty in future responses to CO2 increases (@sfig-ecs;
@fig-methods-dhws; [@grose2023; @shepherd2018]). Projections using Shared Socio-economic
Pathway (SSP) 2-4.5 was used to represent the plausible emission pathway. It represents a
moderate emissions scenario with continued CO2 emissions comparable to present day until
2050, after which emissions gradually decline [@hausfather2020; IPCC -@climate2023]. The
selected GCMs represent 2.5°C to 4.31°C of global atmospheric warming in the 2081-2100
period relative to 1850-1900.

@@

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of positive HotSpot SST Anomaly values, expressed in units of °C-weeks. The
12-week accumulation period reflects the cumulative nature of coral thermal stress. Brief
temperature spikes may be tolerable, but sustained or repeated exposure to elevated
temperatures causes progressively greater physiological stress and bleaching severity. DHW
values integrate both the magnitude and duration of thermal stress: a moderate anomaly
(e.g., 1°C) sustained for 12 weeks produces the same DHW value as a severe anomaly (e.g.,
3°C) lasting 4 weeks, though the biological impacts may differ. DHW estimates for each
indicative reef (see @GBRMPA2022) were then determined by taking the mean DHW value for all
pixels touching each reef outline.

>>>>

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled
1km^2^ resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4. The selected GCMs are a part of the
Sixth Phase of the Coupled Model Intercomparison Project (CMIP6), spanning four 'model
development families' as defined by Brunner et al., [-@brunner2020].

These GCMs were selected based on based on evaluation conducted by Grose et al.,
[-@grose2023], indicating their ability to recreate historical atmospheric temperature and
precipitation conditions in the Australasian region. They span a range of Equilibrium
Climate Sensitivity (ECS) values within the Intergovernmental Panel on Climate Change (IPCC)
assessed likely range. The IPCC defines ECS as the long term warming caused by a doubling of
carbon dioxide above its pre-industrial concentration [@forster2021]. While this assessment
does not consider oceanic data, Grose et al. [-@grose2023] provide a region-specific
selection mechanism which we use as an indicator of general performance of GCMs for the
region. An overview of the selected GCMs are provided in @tbl-gcms. Note here that Grose et
al., [-@grose2023] assessed GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023]
provides data based on GFDL-CM4. Similarly, data for ACCESS-CM2 was used as a stand-in for
ACCESS-ESM1-5. These substitutions are assumed appropriate given they share the same
fundamental representation of the physical climate system.

Projections using Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the
plausible emission pathway. It represents a moderate emissions scenario with continued CO2
emissions comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.31°C **\[TI:
Use K or Celcius?\]** of global atmospheric warming in the 2081-2100 period relative to
1850-1900. The use of this range of GCMs therefore allow for a range of possible climate
futures to be explored, capturing the uncertainty in future responses to CO2 increases
(@sfig-ecs; @fig-methods-dhws; [@grose2023; @shepherd2018]) and enables the assessment of
model projected coral reef behaviour over a broad range of independently derived climate
futures [@abramowitz2019].

@@

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of positive HotSpot SST Anomaly values, expressed in units of °C-weeks. The
12-week accumulation period reflects the cumulative nature of coral thermal stress. Brief
temperature spikes may be tolerable, but sustained or repeated exposure to elevated
temperatures causes progressively greater physiological stress and bleaching severity. DHW
values integrate both the magnitude and duration of thermal stress: a moderate anomaly
(e.g., 1°C) sustained for 12 weeks produces the same DHW value as a severe anomaly (e.g.,
3°C) lasting 4 weeks, though the biological impacts may differ. DHW estimates for each
indicative reef (see @GBRMPA2022) were then determined by taking the mean DHW value for all
pixels touching each reef outline. The process described here was conducted for all 2429
reefs to create DHW projections spanning from 2025 to 2075. Code used to calculate DHW data
and perform analyses involved in this investigation can be found at **\[insert repo
citation\]**.
